---
title: "Chapter 7 - Continuous Probability Distributions"
subtitle: "Uniform, Normal, Exponential, and Related Distributions"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Understand and work with continuous probability distributions
- Calculate probabilities and quantiles for common distributions
- Generate random samples from distributions
- Understand the relationships between distributions
- Apply limit theorems (LLN, CLT, LDP)

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_Prob2.pdf`
:::

## Uniform Distribution

The simplest continuous distribution - all intervals of the same length are equally probable.

### Definition

$$X \sim \text{Uniform}(a, b)$$

$$f(x) = \begin{cases} \frac{1}{b-a} & \text{if } a \leq x \leq b \\ 0 & \text{otherwise} \end{cases}$$

### Properties

$$\mathbb{E}[X] = \frac{a+b}{2}, \quad \text{Var}(X) = \frac{(b-a)^2}{12}$$

### In R

```{r}
# Density
x <- seq(0, 1, length.out = 100)
y <- dunif(x, min = 0, max = 1)
plot(x, y, type = "l", main = "Uniform(0,1) PDF")

# Cumulative distribution
punif(0.5, min = 0, max = 1)  # P(X <= 0.5) = 0.5

# Quantile function
qunif(0.95, min = 0, max = 1)  # 0.95 quantile

# Random generation
samples <- runif(1000, min = -1, max = 1)
hist(samples, breaks = 30, probability = TRUE,
     main = "Histogram of Uniform(-1,1) samples")
lines(density(samples), col = "red")
```

## Normal (Gaussian) Distribution

The most important distribution in statistics - arises from the Central Limit Theorem.

### Definition

$$X \sim \mathcal{N}(\mu, \sigma^2)$$

$$f(x) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right)$$

### Properties

$$\mathbb{E}[X] = \mu, \quad \text{Var}(X) = \sigma^2$$

Skewness = 0, Kurtosis = 3 (excess kurtosis = 0)

### In R

```{r}
# Standard normal
x <- seq(-4, 4, length.out = 200)
y <- dnorm(x, mean = 0, sd = 1)
plot(x, y, type = "l", main = "Standard Normal PDF",
     xlab = "x", ylab = "Density")

# Different parameters
y2 <- dnorm(x, mean = 1, sd = 2)
lines(x, y2, col = "red", lty = 2)
legend("topright", legend = c("N(0,1)", "N(1,4)"),
       col = c("black", "red"), lty = c(1, 2))

# Cumulative distribution
pnorm(0)              # P(Z <= 0) = 0.5
pnorm(1.96)           # ~0.975 (95th percentile)
1 - pnorm(1.96)       # Right tail probability

# Quantile function (critical values)
qnorm(0.025)          # -1.96 (lower 2.5%)
qnorm(0.975)          # 1.96 (upper 2.5%)
qnorm(0.95)           # 1.645 (one-sided 95%)

# Random samples
samples <- rnorm(1000, mean = 100, sd = 15)
hist(samples, breaks = 30, probability = TRUE)
curve(dnorm(x, mean = 100, sd = 15), add = TRUE, col = "red")
```

### Standardization

Any normal can be standardized:

$$Z = \frac{X - \mu}{\sigma} \sim \mathcal{N}(0, 1)$$

```{r}
x <- rnorm(100, mean = 50, sd = 10)
z <- (x - mean(x)) / sd(x)  # Standardize
hist(z, probability = TRUE)
curve(dnorm(x), add = TRUE, col = "red")
```

## Exponential Distribution

Models the time between events in a Poisson process.

### Definition

$$X \sim \text{Exp}(\lambda)$$

$$f(x) = \lambda e^{-\lambda x}, \quad x \geq 0$$

### Properties

$$\mathbb{E}[X] = \frac{1}{\lambda}, \quad \text{Var}(X) = \frac{1}{\lambda^2}$$

Memoryless property: $P(X > s + t | X > s) = P(X > t)$

### In R

```{r}
# Density
x <- seq(0, 5, length.out = 100)
y1 <- dexp(x, rate = 1)
y2 <- dexp(x, rate = 0.5)
y3 <- dexp(x, rate = 2)

plot(x, y1, type = "l", ylab = "Density",
     main = "Exponential Distributions")
lines(x, y2, col = "red", lty = 2)
lines(x, y3, col = "blue", lty = 3)
legend("topright", legend = c("λ=1", "λ=0.5", "λ=2"),
       col = c("black", "red", "blue"), lty = 1:3)

# Mean and variance
lambda <- 2
mean_exp <- 1/lambda
var_exp <- 1/lambda^2
cat("Mean:", mean_exp, "Variance:", var_exp, "\n")

# CDF: P(X <= x)
pexp(1, rate = 1)     # P(X <= 1) with λ=1

# Quantile
qexp(0.5, rate = 1)   # Median
```

## Erlang Distribution

Sum of $n$ independent exponential random variables.

### Definition

$$X \sim \text{Erlang}(n, \lambda)$$

$$f(t) = \frac{\lambda^n t^{n-1}}{(n-1)!} e^{-\lambda t}, \quad t \geq 0$$

### Properties

$$\mathbb{E}[T] = \frac{n}{\lambda}, \quad \text{Var}(T) = \frac{n}{\lambda^2}$$

Special case: $n = 1$ gives the exponential distribution.

### In R

```{r}
# Erlang is a special case of Gamma with integer shape
t <- seq(0, 10, length.out = 100)
plot(t, dgamma(t, shape = 1, rate = 1), type = "l",
     ylab = "Density", main = "Erlang Distributions")
lines(t, dgamma(t, shape = 3, rate = 1), col = "blue")
lines(t, dgamma(t, shape = 5, rate = 1), col = "red")
legend("topright", legend = c("n=1 (Exp)", "n=3", "n=5"),
       col = c("black", "blue", "red"), lty = 1)

# Example: Waiting time until 3rd event
n <- 3
lambda <- 1
mean_erlang <- n/lambda
var_erlang <- n/lambda^2
cat("Erlang(n=3, λ=1): Mean =", mean_erlang, "Variance =", var_erlang, "\n")
```

## Gamma Distribution

Generalizes Erlang by allowing non-integer shape parameter.

### Definition

$$X \sim \text{Gamma}(\alpha, \lambda)$$

$$f(x) = \frac{\lambda^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\lambda x}, \quad x \geq 0$$

where $\Gamma(\alpha) = \int_0^\infty t^{\alpha-1} e^{-t} dt$ is the gamma function.

### Properties

$$\mathbb{E}[X] = \frac{\alpha}{\lambda}, \quad \text{Var}(X) = \frac{\alpha}{\lambda^2}$$

### In R

```{r}
# Different parameterizations in R:
# shape = α, rate = λ (or scale = 1/λ)

x <- seq(0, 15, length.out = 200)
plot(x, dgamma(x, shape = 2, rate = 1), type = "l",
     ylab = "Density", main = "Gamma Distributions")
lines(x, dgamma(x, shape = 2, rate = 0.5), col = "red")
lines(x, dgamma(x, shape = 5, rate = 1), col = "blue")
legend("topright", legend = c("Γ(2,1)", "Γ(2,0.5)", "Γ(5,1)"),
       col = c("black", "red", "blue"), lty = 1)

# Conjugate prior: Gamma is conjugate prior for Poisson rate
```

### Relationships

- $\text{Erlang}(n, \lambda) = \text{Gamma}(n, \lambda)$ when $n$ is integer
- $\chi^2_k = \text{Gamma}(k/2, 1/2)$ (Chi-squared distribution)

## Beta Distribution

Defined on [0,1], useful as a prior distribution for probabilities.

### Definition

$$X \sim \text{Beta}(\alpha, \beta)$$

$$f(x) = \frac{1}{B(\alpha, \beta)} x^{\alpha-1} (1-x)^{\beta-1}, \quad 0 \leq x \leq 1$$

where $B(\alpha, \beta) = \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}$ is the beta function.

### Properties

$$\mathbb{E}[X] = \frac{\alpha}{\alpha + \beta}, \quad \text{Var}(X) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$$

### In R

```{r}
# Plot different beta distributions
x <- seq(0, 1, length.out = 200)
plot(x, dbeta(x, 2, 5), type = "l", ylim = c(0, 3),
     ylab = "Density", main = "Beta Distributions")
lines(x, dbeta(x, 5, 2), col = "red")
lines(x, dbeta(x, 2, 2), col = "blue")
lines(x, dbeta(x, 0.5, 0.5), col = "green")
legend("top", legend = c("Beta(2,5)", "Beta(5,2)", "Beta(2,2)", "Beta(0.5,0.5)"),
       col = c("black", "red", "blue", "green"), lty = 1)

# Conjugate prior: Beta is conjugate prior for Bernoulli/Binomial
# Posterior: Beta(α + k, β + n - k) after observing k successes in n trials
```

## Beta Mixtures

Mixing multiple beta distributions creates multimodal distributions.

```{r}
# Mixture of two betas
x <- seq(0, 1, length.out = 200)
mixture <- 0.5 * dbeta(x, 2, 5) + 0.5 * dbeta(x, 5, 2)
plot(x, mixture, type = "l", ylab = "Density",
     main = "Beta Mixture: 0.5 Beta(2,5) + 0.5 Beta(5,2)")

# Can model complex beliefs that single beta cannot
```

## Limit Theorems

### Law of Large Numbers (LLN)

For i.i.d. random variables $X_1, X_2, \ldots$ with mean $\mu$:

$$\bar{X}_n = \frac{1}{n}\sum_{i=1}^n X_i \xrightarrow{n \to \infty} \mu$$

```{r}
# Demonstration of LLN
set.seed(123)
n <- 10000
x <- runif(n, -1, 1)  # Uniform on [-1,1], true mean = 0
running_mean <- cumsum(x) / (1:n)

plot(running_mean, type = "l", xlab = "n",
     ylab = "Sample Mean",
     main = "Law of Large Numbers")
abline(h = 0, col = "red", lty = 2)
legend("topright", legend = "True mean = 0", col = "red", lty = 2)
```

### Central Limit Theorem (CLT)

For i.i.d. random variables with mean $\mu$ and variance $\sigma^2$:

$$\frac{1}{\sqrt{n}}\sum_{i=1}^n (X_i - \mu) \xrightarrow{d} \mathcal{N}(0, \sigma^2)$$

Equivalently: $\bar{X}_n \approx \mathcal{N}(\mu, \sigma^2/n)$

```{r}
# Demonstration of CLT with uniform distribution
set.seed(42)
n <- 50  # Sample size
m <- 1000  # Number of samples

# Generate sample means from uniform distribution
sample_means <- replicate(m, mean(runif(n, -1, 1)))

# Plot
hist(sample_means, breaks = 30, probability = TRUE,
     main = paste("CLT: Distribution of Sample Mean\nn =", n),
     xlab = "Sample Mean")

# Theoretical normal distribution
theoretical_mean <- 0
theoretical_sd <- sqrt((1/3)/n)  # Var(Uniform(-1,1)) = 1/3
curve(dnorm(x, theoretical_mean, theoretical_sd),
      add = TRUE, col = "red", lwd = 2)
```

### Large Deviation Principle (LDP)

For rare fluctuations far from the mean:

$$P(\bar{X}_n \approx x) \asymp e^{-n\mathcal{I}(x)}$$

where $\mathcal{I}(x)$ is the rate function.

Key idea: Events far from the mean are exponentially rare.

## Unified Picture of Limit Theorems

| Regime | Scaling | Description |
|--------|---------|-------------|
| LLN | $n$ | Deterministic limit |
| CLT | $\sqrt{n}$ | Gaussian fluctuations |
| LDP | $e^{-n}$ | Rare event tails |

**Big picture**: Probability concentrates as data grows.

## Summary

Key takeaways from this chapter:

1. **Uniform**: Equal probability over interval; foundation for random number generation
2. **Normal**: Central to statistics; arises from CLT
3. **Exponential**: Memoryless waiting times; $\lambda$ is rate parameter
4. **Gamma/Erlang**: Sum of exponentials; conjugate prior for Poisson
5. **Beta**: Conjugate prior for Bernoulli/Binomial probabilities
6. **Limit Theorems**: LLN (convergence to mean), CLT (Gaussian fluctuations), LDP (rare events)

## Exercises

::: {.callout-important}
## Exercise 1
Generate 1000 samples from N(100, 225). Calculate the proportion that fall within 1, 2, and 3 standard deviations of the mean. Compare to the empirical rule (68-95-99.7).
:::

::: {.callout-important}
## Exercise 2
If radioactive decay follows an exponential distribution with rate λ = 0.1 per second, what is the probability that a particle decays within the first 5 seconds? What is the expected waiting time?
:::

::: {.callout-important}
## Exercise 3
Verify the CLT with your own simulation using a non-normal distribution (e.g., uniform or exponential).
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_Prob2.pdf`
- "Probability and Statistics for Engineers" by Walpole et al.
- "Statistical Inference" by Casella and Berger
