---
title: "Chapter 1 - Introduction to R for Bayesian Inference"
subtitle: "Getting Started with R: Vectors, Matrices, and Basic Operations"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Install R and RStudio and understand the basic interface
- Create and manipulate vectors and matrices in R
- Perform arithmetic operations on R objects
- Use basic R functions for random number generation
- Understand vectorized operations in R

## Introduction to R

R is a programming language and software environment for statistical computing and graphics. It is widely used among statisticians and data miners for developing statistical software and data analysis.

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_R_first.pdf`
:::

## Getting Started

### Assignment Operator

In R, we use `<-` (or `=`) for assignment:

```{r}
# Assign a value to a variable
x <- 5
x

# Alternative syntax
y = 10
y
```

### Basic Data Types

R has several basic data types:

- **Numeric**: Real numbers (e.g., `3.14`, `-2.5`)
- **Integer**: Whole numbers (e.g., `5L`, `-3L`)
- **Character**: Text strings (e.g., `"hello"`, `'world'`)
- **Logical**: Boolean values (`TRUE`, `FALSE`)

```{r}
# Examples of different types
num_var <- 3.14159
int_var <- 42L
char_var <- "Hello R"
log_var <- TRUE

# Check the type of a variable
class(num_var)
class(int_var)
class(char_var)
class(log_var)
```

## Vectors in R

Vectors are the most basic data structure in R. They are one-dimensional arrays that can hold numeric data, character data, or logical data.

### Creating Vectors

Use the `c()` function to combine values into a vector:

```{r}
# Create a numeric vector
v1 <- c(1, 2, 3, 4, 5)
v1

# Create a character vector
fruits <- c("apple", "banana", "cherry")
fruits

# Create a logical vector
flags <- c(TRUE, FALSE, TRUE, TRUE)
flags

# Sequence of numbers
seq_1_to_10 <- 1:10
seq_1_to_10

# Using seq() function for more control
seq(from = 0, to = 1, by = 0.1)
seq(from = 0, to = 10, length.out = 5)
```

### Vector Operations

R performs element-wise operations on vectors:

```{r}
a <- c(1, 2, 3, 4, 5)
b <- c(10, 20, 30, 40, 50)

# Arithmetic operations
a + b      # Element-wise addition
a - b      # Element-wise subtraction
a * b      # Element-wise multiplication
a / b      # Element-wise division
a^2        # Each element squared

# Vector recycling (shorter vector is repeated)
c <- c(1, 2)
a + c      # c is recycled: c(1, 2, 1, 2, 1)
```

### Vector Functions

```{r}
x <- c(3, 1, 4, 1, 5, 9, 2, 6)

# Basic statistics
length(x)      # Number of elements
sum(x)         # Sum of all elements
mean(x)        # Arithmetic mean
median(x)      # Median value
sd(x)          # Standard deviation
var(x)         # Variance
min(x)         # Minimum value
max(x)         # Maximum value
range(x)       # Min and max
summary(x)     # Summary statistics

# Sorting
sort(x)                # Sort in ascending order
sort(x, decreasing = TRUE)  # Sort in descending order
```

### Accessing Vector Elements

```{r}
v <- c(10, 20, 30, 40, 50)

# Indexing (R uses 1-based indexing!)
v[1]       # First element
v[3]       # Third element
v[c(1, 3, 5)]  # Elements 1, 3, and 5
v[2:4]     # Elements 2 through 4

# Logical indexing
v[v > 25]  # Elements greater than 25
v[v %% 20 == 0]  # Elements divisible by 20

# Exclude elements
v[-1]      # All elements except first
v[-c(2, 4)]  # Exclude elements 2 and 4
```

## Matrices in R

Matrices are two-dimensional arrays where all elements have the same data type.

### Creating Matrices

```{r}
# Create a matrix from a vector
m1 <- matrix(1:12, nrow = 3, ncol = 4)
m1

# By default, matrices are filled by column
# Use byrow = TRUE to fill by row
m2 <- matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
m2

# Create a matrix with specific values
m3 <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3)
m3
```

### Matrix Operations

```{r}
A <- matrix(1:4, nrow = 2)
B <- matrix(c(5, 6, 7, 8), nrow = 2)

# Basic operations
A + B      # Element-wise addition
A - B      # Element-wise subtraction
A * B      # Element-wise multiplication (Hadamard product)
A / B      # Element-wise division

# Matrix multiplication
A %*% B    # True matrix multiplication

# Matrix transpose
t(A)

# Matrix inverse (for square matrices)
C <- matrix(c(4, 7, 2, 6), nrow = 2)
solve(C)   # Inverse of C

# Determinant
det(C)

# Diagonal
diag(A)
diag(3)    # 3x3 identity matrix

# Eigenvalues and eigenvectors
eigen(C)
```

### Accessing Matrix Elements

```{r}
M <- matrix(1:12, nrow = 3, ncol = 4)
M

# Access by row and column
M[1, 2]    # Element in row 1, column 2
M[2, ]     # Entire row 2
M[, 3]     # Entire column 3
M[1:2, 2:3]  # Submatrix (rows 1-2, columns 2-3)

# Dimension of matrix
dim(M)
nrow(M)
ncol(M)
```

## Random Number Generation

R has built-in functions for generating random numbers from various distributions:

```{r}
# Set seed for reproducibility
set.seed(123)

# Uniform distribution
runif(5)              # 5 random numbers between 0 and 1
runif(5, min = -1, max = 1)  # Between -1 and 1

# Normal distribution
rnorm(5)              # Standard normal (mean=0, sd=1)
rnorm(5, mean = 10, sd = 2)   # Normal with mean 10, sd 2

# Other distributions
rbinom(10, size = 20, prob = 0.3)  # Binomial
rpois(10, lambda = 3)   # Poisson
rexp(10, rate = 0.5)    # Exponential

# Plotting random data
x <- rnorm(1000)
hist(x, breaks = 30, main = "Histogram of Standard Normal",
     xlab = "Value", col = "lightblue")
```

## Vectorized Operations

One of Rs most powerful features is vectorization - operations are applied element-wise without explicit loops:

```{r}
# Compare loop vs vectorized approach
x <- 1:1000000

# Loop approach (SLOW)
loop_result <- numeric(length(x))
for (i in 1:length(x)) {
  loop_result[i] <- x[i]^2
}

# Vectorized approach (FAST)
vector_result <- x^2

# They produce the same result
all.equal(loop_result, vector_result)
```

::: {.callout-tip}
## Best Practice
Always use vectorized operations in R instead of loops when possible. Vectorized code is not only more readable but also significantly faster.
:::

## Example: Monte Carlo Estimation of π

We can estimate π by generating random points in a unit square and counting what fraction fall within the unit circle:

```{r}
set.seed(42)
n <- 100000  # Number of random points

# Generate random points in [0,1] x [0,1]
x <- runif(n)
y <- runif(n)

# Check if points are inside unit circle (centered at origin)
# We need points in [0,1], so check distance from (0.5, 0.5) <= 0.5
distances <- sqrt((x - 0.5)^2 + (y - 0.5)^2)
inside_circle <- distances <= 0.5

# Points inside circle / total points = π/4 (for quarter circle)
pi_estimate <- 4 * sum(inside_circle) / n
pi_estimate

# Visualize
plot(x[1:1000], y[1:1000], col = ifelse(inside_circle[1:1000], "blue", "red"),
     pch = ".", asp = 1,
     main = "Monte Carlo Estimation of π")
```

## Summary

Key takeaways from this chapter:

1. R uses `<-` for assignment and is case-sensitive
2. Vectors are created with `c()` and support element-wise operations
3. Matrices are 2D arrays created with `matrix()`
4. R uses 1-based indexing (unlike Python which uses 0-based)
5. Vectorized operations are preferred over loops for performance
6. R has extensive built-in functions for random number generation

## Exercises

::: {.callout-important}
## Exercise 1
Create a vector of the first 20 even numbers, then calculate their mean and standard deviation.
:::

::: {.callout-important}
## Exercise 2
Create a 5×5 identity matrix, then replace the diagonal elements with values 1 through 5.
:::

::: {.callout-important}
## Exercise 3
Generate 1000 random numbers from a normal distribution with mean 50 and standard deviation 10. Calculate what percentage fall within one standard deviation of the mean (i.e., between 40 and 60).
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_R_first.pdf`
- R Documentation: `?matrix`, `?runif`, `?rnorm`
- "R for Data Science" by Hadley Wickham
