---
title: "Chapter 5 - Data Import/Output and Advanced R"
subtitle: "Working with External Data and Advanced R Features"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Import data from various file formats
- Read data from web sources and databases
- Export data in different formats
- Use advanced data structures
- Work with dates and times in R

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_R_4.pdf`
:::

## Reading Data with read.table()

The `read.table()` family of functions provides flexible data import:

```{r, eval=FALSE}
# Basic usage
data <- read.table("filename.txt")

# Common options
data <- read.table("filename.txt",
  header = FALSE,          # First line is NOT column names
  sep = "",                # Whitespace delimiter (default)
  dec = ".",               # Decimal point character
  quote = '""',             # Quoting character
  na.strings = "NA",       # String(s) to interpret as NA
  colClasses = NA,         # Column classes (auto-detect)
  nrows = -1,              # Number of rows to read (-1 = all)
  skip = 0,                # Skip first n lines
  check.names = TRUE,      # Make column names valid R names
  fill = FALSE,            # Fill missing values in short rows
  comment.char = "#"       # Character for comments
)
```

### Specialized Functions

```{r, eval=FALSE}
# CSV (comma-separated values)
read.csv("file.csv")
read.csv2("file.csv")  # European format (semicolon, comma as decimal)

# Delimited files
read.delim("file.txt")     # Tab-delimited
read.delim2("file.txt")    # Tab-delimited with European decimals

# Fixed width format
read.fwf("file.txt", widths = c(5, 10, 8))
```

## Importing Data from Various Sources

### From the Web

```{r, eval=FALSE}
# Read data from URL
url <- "https://raw.githubusercontent.com/.../data.csv"
web_data <- read.csv(url)

# Using httr package for more control
library(httr)
response <- GET("https://api.example.com/data")
content <- content(response, "text")
data <- read.csv(text = content)
```

### From Databases

```{r, eval=FALSE}
# Using RSQLite as example
library(DBI)
library(RSQLite)

# Connect to database
con <- dbConnect(RSQLite::SQLite(), "mydb.sqlite")

# List tables
dbListTables(con)

# Read a table
data <- dbReadTable(con, "mytable")

# Execute SQL query
result <- dbGetQuery(con, "SELECT * FROM mytable WHERE age > 25")

# Close connection
dbDisconnect(con)
```

### Binary Data Formats

```{r, eval=FALSE}
# HDF5 format (large multidimensional arrays)
library(rhdf5)
h5file <- H5Fopen("data.h5")
data <- h5read(h5file, "dataset_name")
H5Fclose(h5file)

# NetCDF format (climate/weather data)
library(ncdf4)
nc <- nc_open("data.nc")
data <- ncvar_get(nc, "variable_name")
nc_close(nc)
```

### Images

```{r, eval=FALSE}
# Using magick package
library(magick)
img <- image_read("path/to/image.png")
print(img)

# For image analysis (pixel values)
library(png)
img_data <- readPNG("image.png")  # Returns numeric array
```

## Exporting Data

### Writing to Files

```{r, eval=FALSE}
# Write CSV
write.csv(data, "output.csv", row.names = FALSE)
write.csv(data, "output.csv", na = "", quote = FALSE)

# Write table (more control)
write.table(data, "output.txt",
  sep = "\t",              # Tab-delimited
  quote = FALSE,           # No quotes
  row.names = FALSE,       # No row names
  col.names = TRUE,        # Include column names
  na = "NA"                # NA representation
)
```

### Saving R Objects

```{r, eval=FALSE}
# RDS format (single object)
saveRDS(data, "data.rds")
data <- readRDS("data.rds")

# RData format (multiple objects)
save(data1, data2, list_obj, file = "workspace.RData")
load("workspace.RData")  # Loads objects into environment

# Save entire workspace
save.image("my_workspace.RData")
```

## Working with Dates and Times

### Date Objects

```{r}
# Create dates
d1 <- as.Date("2024-03-15")
d2 <- as.Date("15/03/2024", format = "%d/%m/%Y")

# Current date
today <- Sys.Date()

# Date arithmetic
d1 + 7           # Add 7 days
d2 - d1          # Difference in days
difftime(d2, d1, units = "weeks")

# Sequence of dates
seq(from = as.Date("2024-01-01"), to = as.Date("2024-12-31"), by = "month")
```

### POSIXct (Calendar Time)

```{r}
# Create datetime
t1 <- as.POSIXct("2024-03-15 14:30:00")
t2 <- as.POSIXct("2024-03-15 14:30:00", tz = "UTC")

# Current time
now <- Sys.time()

# Formatting
format(now, "%Y-%m-%d %H:%M:%S")
format(now, "%A, %B %d, %Y")

# Extract components
as.numeric(format(now, "%Y"))   # Year
as.numeric(format(now, "%m"))   # Month
as.numeric(format(now, "%d"))   # Day
```

### POSIXlt (Local Time)

```{r}
# POSIXlt stores components as a list
t_lt <- as.POSIXlt(Sys.time())
t_lt$hour      # Hour
 t_lt$min       # Minute
t_lt$sec       # Second
t_lt$wday      # Day of week (0-6, Sunday=0)
t_lt$yday      # Day of year (0-365)
```

### Time Durations

```{r}
# Create time difference
one_week <- as.difftime(1, units = "weeks")
two_days <- as.difftime(2, units = "days")

# Add to datetime
future <- Sys.time() + one_week

# lubridate package (tidyverse)
library(lubridate)
now() + days(7) + hours(3)
interval(ymd("2024-01-01"), ymd("2024-12-31")) / days(1)
```

## Set Operations

```{r}
setA <- c("a", "b", "c", "d", "e")
setB <- c("d", "e", "f", "g")

# Union
union(setA, setB)           # All unique elements

# Intersection
intersect(setA, setB)       # Common elements

# Set difference (order matters)
setdiff(setA, setB)         # Elements in A but not in B
setdiff(setB, setA)         # Elements in B but not in A

# Set equality
setequal(setA, setB)        # FALSE

# Membership
setA %in% setB              # Which elements of A are in B
"c" %in% setA               # TRUE
```

## Vector Inspection Functions

```{r}
names <- c("John", "John", "Jim", "Anna", "Beatrix", "Anna")

# Count occurrences
table(names)

# Get unique values
unique(names)

# Find duplicates
duplicated(names)           # TRUE if already appeared

# Remove duplicates
unique(names)
```

## The Tidyverse Ecosystem

### Core Packages

| Package | Purpose |
|---------|---------|
| `ggplot2` | Data visualization |
| `dplyr` | Data manipulation |
| `tidyr` | Data tidying |
| `readr` | Data import |
| `purrr` | Functional programming |
| `tibble` | Modern data frames |
| `stringr` | String manipulation |
| `forcats` | Factor handling |

```{r, eval=FALSE}
# Install tidyverse
install.packages("tidyverse")

# Load core packages
library(tidyverse)
```

### dplyr Basics

```{r, eval=FALSE}
library(dplyr)

# Pipe operator (%>% or |>)
df %>%
  filter(age > 25) %>%
  select(name, score) %>%
  arrange(desc(score)) %>%
  mutate(grade = case_when(
    score >= 90 ~ "A",
    score >= 80 ~ "B",
    TRUE ~ "C"
  ))

# Core verbs
# filter() - select rows
# select() - select columns
# mutate() - create/modify columns
# arrange() - sort rows
# summarise() - aggregate
# group_by() - group data
```

## Practical Examples

### Example 1: Complete Data Pipeline

```{r, eval=FALSE}
# Load and process data
library(tidyverse)

# Read data
data <- read_csv("students.csv", na = c("", "NA", "N/A"))

# Clean and transform
clean_data <- data %>%
  filter(!is.na(age)) %>%
  mutate(
    age_group = cut(age, breaks = c(0, 20, 25, 30, Inf),
                   labels = c("<20", "20-25", "25-30", "30+")),
    gpa_numeric = as.numeric(gsub("[^0-9.]", "", gpa)),
    enrollment_date = as.Date(enrollment_date, format = "%m/%d/%Y")
  ) %>%
  select(-raw_score)  # Remove unnecessary column

# Save cleaned data
write_csv(clean_data, "students_clean.csv")
```

### Example 2: Working with Multiple Files

```{r, eval=FALSE}
# Read and combine multiple CSV files
library(tidyverse)

# Get list of files
files <- list.files("data/", pattern = "\\.csv$", full.names = TRUE)

# Read and combine all files
combined_data <- files %>%
  map_dfr(read_csv, .id = "file_id")

# Alternative using base R
combined <- do.call(rbind, lapply(files, read.csv))
```

## Summary

Key takeaways from this chapter:

1. **Data Import**: `read.table()` family handles most text formats; specialized packages for binary formats
2. **Web Data**: Can read directly from URLs; use httr for API access
3. **Databases**: DBI package provides unified interface to SQL databases
4. **Dates/Times**: Use `as.Date()` for dates, `POSIXct` for timestamps, `lubridate` for easier handling
5. **Tidyverse**: Opinionated collection of packages for modern data science

## Exercises

::: {.callout-important}
## Exercise 1
Write code to read a CSV file from a URL, convert date columns to proper Date format, and save the cleaned data as an RDS file.
:::

::: {.callout-important}
## Exercise 2
Create a function that takes two data frames and returns information about which rows are common, which are only in the first, and which are only in the second.
:::

::: {.callout-important}
## Exercise 3
Working with dates: Calculate the number of weekdays between two dates (excluding weekends).
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_R_4.pdf`
- R Documentation: `?read.table`, `?POSIXct`, `?Date`
- "R for Data Science" by Hadley Wickham
