---
title: "Chapter 8 - Discrete Probability Distributions"
subtitle: "Binomial, Poisson, Geometric, and Other Discrete Distributions"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Understand fundamental discrete probability distributions
- Calculate probabilities for common discrete scenarios
- Model count data and waiting times
- Use R functions for discrete distributions
- Solve problems involving combinations and permutations

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_Discrete_Prob.pdf`
:::

## Combinatorics: Permutations and Combinations

### Permutations

Number of ways to arrange $r$ objects from $n$ distinct objects where order matters:

$$P(n, r) = \frac{n!}{(n-r)!}$$

### Combinations

Number of ways to choose $r$ objects from $n$ distinct objects where order doesnt matter:

$$\binom{n}{r} = \frac{n!}{r!(n-r)!}$$

### In R

```{r}
# Factorial
factorial(5)   # 5! = 120

# Combinations
choose(10, 3)  # C(10,3) = 120

# Permutations
perm <- function(n, r) factorial(n) / factorial(n - r)
perm(10, 3)    # P(10,3) = 720

# Practical example: How many ways to choose 3 people from 10?
choose(10, 3)  # 120
```

## Bernoulli Distribution

The simplest discrete distribution - a single trial with two outcomes.

### Definition

$$X \sim \text{Bernoulli}(p)$$

$$P(X = k) = \begin{cases} p & \text{if } k = 1 \text{ (success)} \\ 1-p & \text{if } k = 0 \text{ (failure)} \end{cases}$$

### Properties

$$\mathbb{E}[X] = p, \quad \text{Var}(X) = p(1-p)$$

## Binomial Distribution

Models the number of successes in $n$ independent Bernoulli trials.

### Definition

$$X \sim \text{Binomial}(n, p)$$

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}, \quad k = 0, 1, \ldots, n$$

### Properties

$$\mathbb{E}[X] = np, \quad \text{Var}(X) = np(1-p)$$

### In R

```{r}
# Probability mass function: P(X = k)
dbinom(3, size = 10, prob = 0.4)  # P(X=3) with n=10, p=0.4

# Cumulative distribution: P(X <= k)
pbinom(3, size = 10, prob = 0.4)  # P(X <= 3)
pbinom(3, size = 10, prob = 0.4, lower.tail = FALSE)  # P(X > 3)

# Quantile function
qbinom(0.95, size = 10, prob = 0.4)  # 95th percentile

# Random generation
rbinom(100, size = 10, prob = 0.4)

# Visualization
x <- 0:10
y <- dbinom(x, size = 10, prob = 0.4)
barplot(y, names.arg = x, xlab = "k", ylab = "P(X=k)",
        main = "Binomial(10, 0.4)", col = "steelblue")
```

### Example: Restaurant Problem

```{r}
# In a restaurant, 8 entrees are fish, 12 are beef, 10 are poultry (total 30)
# What is the probability that 2 of the next 4 customers order fish?

# X ~ Binomial(n=4, p=8/30)
n <- 4
p <- 8/30
prob <- dbinom(2, size = n, prob = p)
cat("P(X = 2) =", prob, "\n")

# Exact calculation
dbinom(2, size = 4, prob = 8/30)
```

## Multinomial Distribution

Generalization of binomial to $k$ possible outcomes.

### Definition

$$P(X_1 = x_1, \ldots, X_k = x_k) = \frac{n!}{x_1! \cdots x_k!} p_1^{x_1} \cdots p_k^{x_k}$$

where $\sum_{i=1}^k x_i = n$ and $\sum_{i=1}^k p_i = 1$.

### In R

```{r}
# Generate multinomial samples
rmultinom(n = 1, size = 20, prob = c(0.2, 0.5, 0.3))

# Example: 20 trials, 3 categories with probabilities 0.2, 0.5, 0.3
set.seed(123)
rmultinom(n = 5, size = 20, prob = c(0.2, 0.5, 0.3))
```

## Geometric Distribution

Models the number of trials until the first success.

### Definition

$$P(N = n) = (1-p)^{n-1}p, \quad n = 1, 2, \ldots$$

### Properties

$$\mathbb{E}[N] = \frac{1}{p}, \quad \text{Var}(N) = \frac{1-p}{p^2}$$

**Memoryless Property**: $P(N > n + m | N > n) = P(N > m)$

### In R

```{r}
# Note: R uses number of failures BEFORE first success!
# So P(N_fail = k) = (1-p)^k * p

# P(first success at trial 5)
dgeom(4, prob = 0.3)  # 4 failures before 1st success

# Alternative: P(N = 5) directly
p_geom <- function(n, p) (1-p)^(n-1) * p
p_geom(5, 0.3)

# Mean and variance
mean_geom <- function(p) 1/p
var_geom <- function(p) (1-p)/p^2
cat("Geometric(p=0.3): Mean =", mean_geom(0.3), "Variance =", var_geom(0.3), "\n")

# Sampling waiting times
rgeom(1000, prob = 0.3)  # Number of failures before success
hist(rgeom(10000, prob = 0.3), breaks = 30,
     main = "Geometric Distribution (failures before success)",
     xlab = "Number of Failures")
```

## Negative Binomial Distribution

Models the number of failures before the $r$-th success.

### Definition

$$P(X = k) = \binom{k+r-1}{k} (1-p)^k p^r, \quad k = 0, 1, 2, \ldots$$

### Properties

$$\mathbb{E}[X] = \frac{r(1-p)}{p}, \quad \text{Var}(X) = \frac{r(1-p)}{p^2}$$

Special case: $r = 1$ gives the geometric distribution.

### In R

```{r}
# Plot negative binomial distributions
x <- 0:30
y1 <- dnbinom(x, size = 1, prob = 0.3)  # Geometric
y2 <- dnbinom(x, size = 5, prob = 0.3)

plot(x, y1, type = "h", ylim = c(0, max(y1, y2)),
     xlab = "k", ylab = "P(X=k)", main = "Negative Binomial")
lines(x + 0.15, y2, type = "h", col = "red")
legend("topright", legend = c("r=1 (Geometric)", "r=5"),
       col = c("black", "red"), lty = 1)

# Example: Expected number of failures before 3rd success with p=0.3
r <- 3
p <- 0.3
mean_nb <- r*(1-p)/p
cat("Mean failures before", r, "th success:", mean_nb, "\n")
```

## Hypergeometric Distribution

Models sampling **without replacement** from a finite population.

### Definition

Draw $n$ items from a population of size $N$ containing $K$ "success" items.

$$P(X = k) = \frac{\binom{K}{k} \binom{N-K}{n-k}}{\binom{N}{n}}$$

### Properties

$$\mathbb{E}[X] = n\frac{K}{N}, \quad \text{Var}(X) = n\frac{K}{N}\left(1-\frac{K}{N}\right)\frac{N-n}{N-1}$$

The factor $\frac{N-n}{N-1}$ is the finite population correction.

### In R

```{r}
# Example: Urn with 40 white and 60 black balls (100 total)
# Draw 20 balls. Probability that at least 50 black balls remain?
# (i.e., at most 10 black balls drawn)

N <- 100    # Total balls
K <- 60     # Black balls (successes)
n <- 20     # Draws

# P(X <= 10) where X ~ Hypergeometric(N=100, K=60, n=20)
prob <- phyper(10, m = K, n = N - K, k = n)
cat("P(at most 10 black balls drawn) =", prob, "\n")

# Compare to binomial approximation
pbinom(10, size = 20, prob = 0.6)  # Approximation
```

## Poisson Distribution

Models the number of rare, independent events in a fixed interval.

### Definition

$$X \sim \text{Poisson}(\lambda)$$

$$P(X = k) = \frac{\lambda^k e^{-\lambda}}{k!}, \quad k = 0, 1, 2, \ldots$$

### Properties

$$\mathbb{E}[X] = \text{Var}(X) = \lambda$$

### In R

```{r}
# Probability mass function
dpois(3, lambda = 2)  # P(X=3) with λ=2

# Cumulative distribution
ppois(3, lambda = 2)  # P(X <= 3)

# Quantile
qpois(0.95, lambda = 2)  # 95th percentile

# Random generation
rpois(100, lambda = 2)

# Visualization
x <- 0:15
y1 <- dpois(x, lambda = 2)
y2 <- dpois(x, lambda = 5)
y3 <- dpois(x, lambda = 10)

plot(x, y1, type = "h", ylim = c(0, 0.35),
     xlab = "k", ylab = "P(X=k)", main = "Poisson Distributions")
lines(x + 0.15, y2, type = "h", col = "blue")
lines(x + 0.3, y3, type = "h", col = "red")
legend("topright", legend = c("λ=2", "λ=5", "λ=10"),
       col = c("black", "blue", "red"), lty = 1)
```

### Poisson as Limit of Binomial

The Poisson distribution is the limit of Binomial$(n, p)$ as $n \to \infty$ and $p \to 0$ with $np \to \lambda$.

```{r}
# Demonstration: Binomial approaches Poisson
n <- 100
p <- 0.05
lambda <- n * p

x <- 0:20
binom_probs <- dbinom(x, size = n, prob = p)
pois_probs <- dpois(x, lambda = lambda)

plot(x, binom_probs, type = "h", col = "blue", lwd = 2,
     xlab = "k", ylab = "Probability",
     main = paste("Binomial(n=", n, ", p=", p,") vs Poisson(λ=", lambda, ")"))
lines(x + 0.15, pois_probs, type = "h", col = "red", lwd = 2)
legend("topright", legend = c("Binomial", "Poisson"),
       col = c("blue", "red"), lty = 1)
```

## Relationships Between Discrete Distributions

The discrete distributions form a hierarchy:

| Distribution | Scenario |
|--------------|----------|
| **Bernoulli** | Single trial |
| **Binomial** | Count of successes in $n$ trials |
| **Multinomial** | Multiple outcome categories |
| **Geometric** | Waiting time for 1st success |
| **Negative Binomial** | Waiting time for $r$-th success |
| **Hypergeometric** | Sampling without replacement |
| **Poisson** | Rare events in continuous time/space |

## Summary Table

| Distribution | Mean | Variance |
|------------|------|----------|
| Bernoulli($p$) | $p$ | $p(1-p)$ |
| Binomial($n, p$) | $np$ | $np(1-p)$ |
| Geometric($p$) | $1/p$ | $(1-p)/p^2$ |
| Neg. Binomial($r, p$) | $r(1-p)/p$ | $r(1-p)/p^2$ |
| Hypergeometric($N, K, n$) | $nK/N$ | $nK/N(1-K/N)(N-n)/(N-1)$ |
| Poisson($\lambda$) | $\lambda$ | $\lambda$ |

## Summary

Key takeaways from this chapter:

1. **Combinatorics**: $\binom{n}{r}$ counts ways to choose; $\frac{n!}{(n-r)!}$ counts arrangements
2. **Binomial**: Count of successes in $n$ independent trials
3. **Poisson**: Rare events; single parameter $\lambda$ equals mean and variance
4. **Geometric**: Memoryless waiting time for first success
5. **Hypergeometric**: Sampling without replacement (finite population correction)

## Exercises

::: {.callout-important}
## Exercise 1
A fair die is rolled 10 times. What is the probability of getting exactly two 6s? At least one 6?
:::

::: {.callout-important}
## Exercise 2
If phone calls arrive at a call center according to a Poisson process with rate 2 per minute, what is the probability of receiving exactly 5 calls in 2 minutes? What is the expected number of calls?
:::

::: {.callout-important}
## Exercise 3
A deck of 52 cards contains 13 hearts. If you draw 5 cards without replacement, what is the probability of getting exactly 2 hearts? Compare to the binomial approximation (with replacement).
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_Discrete_Prob.pdf`
- "Introduction to Probability" by Blitzstein and Hwang
- "Probability: Theory and Examples" by Durrett
