---
title: "Chapter 6 - Graphics in R"
subtitle: "Data Visualization with Base R and ggplot2"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Create various plots using base R graphics
- Use ggplot2 for advanced visualizations
- Customize plot appearance
- Save plots to files
- Create publication-quality figures

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_R_graphics.pdf`
:::

## Base R Graphics

### Basic Plot Function

The `plot()` function is the foundation of R graphics:

```{r}
# Simple scatter plot
x <- 1:10
y <- x^2
plot(x, y)

# Customized plot
plot(x, y,
  main = "Quadratic Function",    # Title
  xlab = "x values",              # X-axis label
  ylab = "y values",              # Y-axis label
  pch = 19,                       # Point character (solid circle)
  col = "blue",                   # Color
  cex = 1.5                       # Point size
)
```

### Plot Types

```{r}
# Line plot
plot(x, y, type = "l", col = "red")

# Both points and lines
plot(x, y, type = "b", pch = 19)

# Common type values:
# "p" = points (default)
# "l" = lines
# "b" = both
# "o" = overplotted
# "h" = histogram-like
# "s" = stairs
# "n" = no plotting (just axes)
```

### Point Characters and Colors

```{r}
# All point types
plot(1:25, rep(1, 25), pch = 1:25, cex = 2)
text(1:25, rep(0.8, 25), labels = 1:25)

# Colors
plot(x, y, col = "red")           # Named color
plot(x, y, col = "#FF5733")       # Hex color
plot(x, y, col = 2)               # Color index

# Color palettes
colors <- rainbow(10)
plot(1:10, col = colors, pch = 19, cex = 3)
```

### Adding Elements to Plots

```{r}
# Base plot
x <- seq(-3, 3, length.out = 100)
y <- dnorm(x)
plot(x, y, type = "l", main = "Standard Normal Distribution")

# Add points
points(0, dnorm(0), pch = 19, col = "red", cex = 2)

# Add lines
abline(v = 0, col = "gray", lty = 2)  # Vertical line
abline(h = 0, col = "gray", lty = 2)  # Horizontal line

# Add text
text(0, dnorm(0) + 0.02, "Mean", pos = 3)

# Add legend
legend("topright",
       legend = c("Density", "Mean"),
       col = c("black", "red"),
       lty = c(1, NA),
       pch = c(NA, 19))

# Add arrows
arrows(-2, 0.2, -1, dnorm(-1), length = 0.1)
```

### Histograms and Density Plots

```{r}
# Generate data
set.seed(123)
data <- rnorm(1000, mean = 50, sd = 10)

# Histogram
hist(data, breaks = 30,
     main = "Distribution of Values",
     xlab = "Value",
     col = "lightblue",
     border = "white")

# Add density curve
hist(data, breaks = 30, probability = TRUE,
     main = "Histogram with Density",
     xlab = "Value",
     col = "lightblue")
lines(density(data), col = "red", lwd = 2)
```

### Box Plots

```{r}
# Generate data
group1 <- rnorm(100, mean = 50, sd = 10)
group2 <- rnorm(100, mean = 55, sd = 8)
group3 <- rnorm(100, mean = 45, sd = 12)

# Create box plot
boxplot(list(Group1 = group1, Group2 = group2, Group3 = group3),
        main = "Comparison of Groups",
        ylab = "Value",
        col = c("red", "green", "blue"))

# Horizontal box plot
boxplot(group1, group2, group3,
        names = c("A", "B", "C"),
        horizontal = TRUE,
        col = "lightgreen")
```

### Bar Plots

```{r}
# Simple bar plot
counts <- c(12, 8, 15, 6)
names(counts) <- c("A", "B", "C", "D")
barplot(counts, col = "steelblue")

# Grouped bar plot
data <- matrix(c(12, 8, 15, 6, 9, 11), nrow = 2)
barplot(data,
        beside = TRUE,
        col = c("darkblue", "red"),
        legend.text = c("Group 1", "Group 2"))
```

### Multiple Plots

```{r}
# Layout multiple plots
par(mfrow = c(2, 2))  # 2 rows, 2 columns

plot(rnorm(100), main = "Plot 1")
plot(runif(100), main = "Plot 2")
hist(rnorm(100), main = "Plot 3")
boxplot(rnorm(100), main = "Plot 4")

par(mfrow = c(1, 1))  # Reset to single plot
```

## ggplot2 Graphics

ggplot2 is a powerful visualization system based on the grammar of graphics.

### Basic Structure

```{r}
library(ggplot2)

# Sample data
df <- data.frame(
  x = 1:10,
  y = (1:10)^2,
  group = rep(c("A", "B"), each = 5)
)

# Basic ggplot
ggplot(df, aes(x = x, y = y)) +
  geom_point()

# Customized plot
ggplot(df, aes(x = x, y = y, color = group)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Quadratic Growth",
       x = "Input",
       y = "Output") +
  theme_minimal()
```

### Geometric Objects (Geoms)

```{r}
# Scatter plot
ggplot(df, aes(x, y)) + geom_point()

# Line plot
ggplot(df, aes(x, y)) + geom_line()

# Bar plot
df_cat <- data.frame(category = c("A", "B", "C"), value = c(10, 20, 15))
ggplot(df_cat, aes(category, value)) + geom_bar(stat = "identity")

# Histogram
ggplot(data.frame(x = rnorm(1000)), aes(x)) + geom_histogram(bins = 30)

# Density plot
ggplot(data.frame(x = rnorm(1000)), aes(x)) + geom_density()

# Box plot
ggplot(data.frame(group = rep(c("A", "B"), each = 50),
                  value = c(rnorm(50), rnorm(50, mean = 2))),
       aes(group, value)) + geom_boxplot()

# Smooth line with confidence interval
ggplot(df, aes(x, y)) + geom_point() + geom_smooth()
```

### Aesthetics

```{r}
# Create sample data
data <- data.frame(
  x = rep(1:10, 2),
  y = c(1:10, (1:10)^1.2),
  group = rep(c("Linear", "Power"), each = 10),
  size = rep(c(2, 4), each = 10)
)

# Aesthetic mappings
ggplot(data, aes(x = x, y = y)) +
  geom_point(aes(color = group, size = size)) +
  geom_line(aes(linetype = group))

# Manual scales
ggplot(data, aes(x, y, color = group)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("red", "blue"))
```

### Facets

```{r}
# Create faceted data
facet_data <- data.frame(
  x = rep(1:10, 6),
  y = c(rep(1:10, 3), rep((1:10)^1.5, 3)),
  group = rep(c("A", "B", "C"), each = 20),
  type = rep(c("Linear", "Power"), each = 10, times = 3)
)

# Facet by one variable
ggplot(facet_data, aes(x, y)) +
  geom_point() +
  facet_wrap(~ group)

# Facet by two variables
ggplot(facet_data, aes(x, y, color = group)) +
  geom_point() +
  facet_grid(type ~ group)
```

### Themes

```{r}
p <- ggplot(df, aes(x, y)) + geom_point()

# Built-in themes
p + theme_gray()      # Default
p + theme_bw()        # Black and white
p + theme_minimal()   # Minimal
p + theme_classic()   # Classic

# Custom theme
p + theme(
  plot.title = element_text(size = 20, face = "bold"),
  axis.title = element_text(size = 14),
  axis.text = element_text(size = 12),
  panel.background = element_rect(fill = "lightgray"),
  panel.grid.major = element_line(color = "white")
)
```

## Saving Plots

```{r, eval=FALSE}
# Base R - save current plot
png("plot.png", width = 800, height = 600)
plot(1:10, (1:10)^2)
dev.off()

# Multiple formats
pdf("plot.pdf", width = 8, height = 6)
plot(1:10, (1:10)^2)
dev.off()

# SVG (scalable)
svg("plot.svg", width = 8, height = 6)
plot(1:10, (1:10)^2)
dev.off()

# ggplot2 - ggsave (auto-detects format)
ggsave("myplot.png", plot = p, width = 8, height = 6, dpi = 300)
ggsave("myplot.pdf", plot = p, width = 8, height = 6)
```

## Summary

Key takeaways from this chapter:

1. **Base R**: `plot()` is flexible but complex; good for quick exploration
2. **ggplot2**: Grammar of graphics approach; better for publication-quality plots
3. **Aesthetics**: Map variables to visual properties (color, size, shape)
4. **Geoms**: Different plot types (point, line, bar, histogram, etc.)
5. **Themes**: Control appearance with pre-built or custom themes

## Exercises

::: {.callout-important}
## Exercise 1
Create a scatter plot showing the relationship between two variables. Add a trend line, color points by a third variable, and save as a high-resolution PNG.
:::

::: {.callout-important}
## Exercise 2
Create a histogram with a density curve overlaid. Use proper labels and a nice color scheme.
:::

::: {.callout-important}
## Exercise 3
Using ggplot2, create a faceted plot showing data for different groups side by side.
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_R_graphics.pdf`
- "ggplot2: Elegant Graphics for Data Analysis" by Hadley Wickham
- R Graphics Cookbook by Winston Chang
