---
title: "Chapter 4 - Lists and Data Frames in R"
subtitle: "Working with Complex Data Structures"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Create and manipulate lists in R
- Build and modify data frames
- Understand subsetting of data frames
- Import and export data
- Work with tibbles (modern data frames)

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_R_data.pdf`
:::

## Lists in R

Lists are Rs most flexible data structure - they can contain elements of different types and lengths.

### Creating Lists

```{r}
# Create a list with different types
my_list <- list(
  name = "Alice",
  age = 25,
  scores = c(85, 90, 78),
  passed = TRUE
)
my_list

# Empty list
empty_list <- list()

# List from vector (each element becomes list element)
vec <- 1:5
as.list(vec)
```

### Accessing List Elements

```{r}
person <- list(name = "Bob", age = 30, scores = c(70, 85, 90))

# Access by index (returns a list)
person[1]           # List with one element
class(person[1])

# Access by name (returns a list)
person["age"]       # List with one element

# Access with $ (returns the element directly)
person$name         # "Bob"
person$age          # 30

# Access with [[ ]] (returns the element directly)
person[[1]]         # "Bob"
person[["scores"]]  # c(70, 85, 90)

# Difference between [ ] and [[ ]]
person[1:2]         # List with 2 elements
# person[[1:2]]     # ERROR! [[ ]] accepts only single element
```

### Modifying Lists

```{r}
# Add new element
person$city <- "New York"
person

# Modify existing element
person$age <- 31
person[["age"]] <- 32

# Remove element
person$city <- NULL
person

# Combine lists
list1 <- list(a = 1, b = 2)
list2 <- list(c = 3, d = 4)
combined <- c(list1, list2)
combined
```

### List Operations

```{r}
# Length of list
length(person)

# Names of list elements
names(person)

# Check if element exists
"age" %in% names(person)

# Apply function to each element
lapply(person, class)

# Recursive unlisting
nested <- list(a = 1, b = list(c = 2, d = 3))
unlist(nested)  # Flattens to vector
unlist(nested, recursive = FALSE)
```

## Data Frames in R

Data frames are the workhorse of R data analysis - two-dimensional tables where columns can be different types.

### Creating Data Frames

```{r}
# Create a data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie", "Diana"),
  age = c(25, 30, 35, 28),
  score = c(85.5, 92.0, 78.5, 88.0),
  passed = c(TRUE, TRUE, FALSE, TRUE)
)
df

# Structure of data frame
str(df)

# Summary statistics
summary(df)

# Dimensions
nrow(df)
ncol(df)
dim(df)

# Column and row names
names(df)       # Column names
colnames(df)    # Same as names()
rownames(df)    # Row names (usually "1", "2", "3", ...)
```

### Accessing Data Frame Elements

```{r}
# Access columns (all return vectors)
df$age          # Vector
df[["age"]]     # Vector
df[[2]]         # Vector (second column)
df[, 2]         # Vector (second column)

# Access columns (returns data frame)
df["age"]       # Data frame with one column
df[2]           # Data frame with second column
df[, 2, drop = FALSE]  # Data frame (preserves structure)

# Access rows
df[1, ]         # First row (as data frame)
df[1:2, ]       # First two rows

# Access specific elements
df[1, 2]        # Row 1, column 2
df$age[1]       # First element of age column
df[1, "age"]    # Row 1, age column
```

### Subsetting Data Frames

```{r}
# Logical subsetting
df[df$age > 25, ]           # Rows where age > 25
df[df$passed == TRUE, ]       # Rows where passed is TRUE

# Select specific columns
df[, c("name", "score")]     # Only name and score columns
df[1:2, 2:3]                  # First 2 rows, columns 2-3

# subset() function
subset(df, age > 25)                     # Rows where age > 25
subset(df, age > 25, select = c(name, score))  # Select columns too

# Which rows meet condition
which(df$age > 25)           # Returns indices
```

### Modifying Data Frames

```{r}
# Add new column
df$grade <- c("B", "A", "C", "B+")
df$passed_10 <- df$score >= 80
df

# Modify existing column
df$age <- df$age + 1  # Everyone is 1 year older

# Remove column
df$grade <- NULL

# Rename columns
names(df)[names(df) == "score"] <- "final_score"
names(df)

# Reorder columns
df <- df[, c("name", "age", "final_score", "passed", "passed_10")]
df
```

## Importing and Exporting Data

### Reading Data

```{r, eval=FALSE}
# Read CSV file
data <- read.csv("data.csv")
data <- read.csv("data.csv", header = TRUE, stringsAsFactors = FALSE)

# Read from URL
url_data <- read.csv("https://example.com/data.csv")

# Read with specific options
data <- read.csv("data.csv",
               header = TRUE,          # First row is column names
               sep = ",",              # Field separator
               na.strings = c("NA", ""),  # What to treat as NA
               stringsAsFactors = FALSE)  # Don't convert strings to factors

# Read table (general delimiter)
data <- read.table("data.txt", header = TRUE, sep = "\t")

# Read RDS (R serialized format)
data <- readRDS("data.rds")

# Read Excel (requires readxl package)
# library(readxl)
# data <- read_excel("data.xlsx", sheet = "Sheet1")
```

### Writing Data

```{r, eval=FALSE}
# Write to CSV
write.csv(df, "output.csv", row.names = FALSE)
write.csv(df, "output.csv", quote = FALSE, na = "")

# Write to RDS (preserves R objects exactly)
saveRDS(df, "data.rds")

# Save multiple objects
save(df, list, file = "workspace.RData")

# Write to table
write.table(df, "output.txt", sep = "\t", row.names = FALSE)
```

## The Tidyverse: Tibbles

Tibbles are modern data frames with improved printing and behavior.

### Creating Tibbles

```{r}
library(tibble)

# Create a tibble
df_tibble <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  score = c(85.5, 92.0, 78.5)
)
df_tibble

# Convert data frame to tibble
df_as_tibble <- as_tibble(df)
df_as_tibble

# tribble: transposed tibble (row-wise entry)
df_tribble <- tribble(
  ~name,    ~age, ~score,
  "Alice",   25,   85.5,
  "Bob",     30,   92.0,
  "Charlie", 35,   78.5
)
df_tribble
```

### Tibble Features

```{r}
# Tibbles don't convert strings to factors
class(df_tibble$name)  # "character"

# Tibbles have nicer printing
# Only shows first 10 rows
# Shows column types
# Doesn't print row names

# Tibbles are stricter about subsetting
df_tibble$name         # Works (returns vector)
df_tibble[, "age"]    # Returns tibble (not vector!)
df_tibble[["age"]]     # Returns vector
```

### List Columns in Tibbles

```{r}
# Tibbles can have list columns (advanced feature)
df_with_list <- tibble(
  id = 1:3,
  name = c("A", "B", "C"),
  measurements = list(
    c(1.2, 2.3, 3.4),
    c(2.1, 3.2),
    c(4.5, 5.6, 6.7, 7.8)
  )
)
df_with_list

# Access list column
df_with_list$measurements[[1]]  # First person's measurements
```

## The data.table Package

For large datasets, data.table offers faster operations.

```{r}
library(data.table)

# Create data.table
dt <- data.table(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  score = c(85.5, 92.0, 78.5)
)
dt

# Fast subsetting
dt[age > 25]              # Filter rows
dt[, .(name, score)]      # Select columns
dt[, mean_score := mean(score)]  # Add column
dt[, .(avg_age = mean(age)), by = score > 80]  # Group by
```

::: {.callout-tip}
## Tidyverse vs data.table
- **Tidyverse** (dplyr/tibble): Readable code, great for analysis
- **data.table**: Fast performance, memory efficient
- Both are excellent; choose based on your needs
:::

## Practical Examples

### Example 1: Creating a Dataset

```{r}
# Create sample dataset
set.seed(123)
students <- data.frame(
  id = 1:100,
  gender = sample(c("M", "F"), 100, replace = TRUE),
  age = round(rnorm(100, mean = 20, sd = 2)),
  math_score = round(rnorm(100, mean = 75, sd = 10)),
  english_score = round(rnorm(100, mean = 70, sd = 12)),
  stringsAsFactors = FALSE
)

# Add derived columns
students$total_score <- students$math_score + students$english_score
students$passed <- students$total_score >= 140

# Summary
head(students)
summary(students)
```

### Example 2: Data Cleaning

```{r}
# Handle missing values
df_with_na <- data.frame(
  a = c(1, 2, NA, 4),
  b = c(5, NA, 7, 8),
  c = c(9, 10, 11, 12)
)

# Remove rows with any NA
df_no_na <- na.omit(df_with_na)
df_no_na

# Replace NAs with specific value
df_filled <- df_with_na
df_filled$a[is.na(df_filled$a)] <- mean(df_filled$a, na.rm = TRUE)
```

## Summary

Key takeaways from this chapter:

1. **Lists**: Flexible containers for mixed data types; use [[ ]] or $ to access elements
2. **Data Frames**: Two-dimensional data structures; columns can be different types
3. **Subsetting**: Use [row, column] syntax; logical conditions for filtering
4. **Import/Export**: read.csv/write.csv for common formats; saveRDS for R objects
5. **Tibbles**: Modern data frames with better printing and stricter behavior

## Exercises

::: {.callout-important}
## Exercise 1
Create a data frame with student information (name, age, major, GPA). Add a column that classifies each student as "Honors" (GPA >= 3.5), "Good" (GPA >= 3.0), or "Average".
:::

::: {.callout-important}
## Exercise 2
Write a function that takes a data frame and returns a summary of missing values by column.
:::

::: {.callout-important}
## Exercise 3
Create a list containing data frames for three different classes. Write code to calculate the average score for each class and return a summary data frame.
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_R_data.pdf`
- R Documentation: `?data.frame`, `?list`, `?tibble`
- "R for Data Science" by Hadley Wickham
