---
title: "Chapter 3 - Data Types and Precision in R"
subtitle: "Understanding R Data Types and Floating Point Arithmetic"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Understand the different atomic data types in R
- Recognize issues with floating point precision
- Use type coercion appropriately
- Handle special values (NA, NaN, Inf) correctly

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_R_third.pdf`
:::

## Atomic Data Types in R

R has six basic atomic vector types:

| Type | Mode | Description |
|------|------|-------------|
| `logical` | logical | TRUE/FALSE values |
| `integer` | numeric | Whole numbers (L suffix) |
| `double` | numeric | Real numbers |
| `character` | character | Text strings |
| `complex` | complex | Complex numbers |
| `raw` | raw | Raw bytes |

### Logical Type

```{r}
# Logical values
l1 <- TRUE
l2 <- FALSE

# Logical operations
!l1           # Negation
l1 & l2       # AND
l1 | l2       # OR
xor(l1, l2)   # Exclusive OR

# Comparison operators return logical
x <- 5
x < 10       # TRUE
x >= 5       # TRUE
x == 5        # TRUE
x != 5        # FALSE

# Any and All
any(c(FALSE, TRUE, FALSE))   # TRUE
all(c(TRUE, TRUE, TRUE))       # TRUE
```

### Numeric Types: Integer and Double

```{r}
# Integer (explicit with L suffix)
int_var <- 42L
class(int_var)
typeof(int_var)

# Double (default for numbers with decimal point)
dbl_var <- 42.0
class(dbl_var)
typeof(dbl_var)

# Even integers stored as double by default
num <- 42
class(num)    # "numeric" (which means double)

# Converting between types
as.integer(3.14)    # 3 (truncates)
as.double(42L)      # 42.0
```

### Character Type

```{r}
# Character strings
str1 <- "Hello"
str2 <- 'World'  # Single quotes also work

# String concatenation
paste(str1, str2)           # "Hello World"
paste0(str1, str2)          # "HelloWorld"

# String manipulation
toupper("hello")            # "HELLO"
tolower("WORLD")            # "world"
substr("Hello World", 1, 5) # "Hello"
nchar("Hello")              # 5

# Converting to character
as.character(42)            # "42"
as.character(3.14159)       # "3.14159"
```

### Complex Numbers

```{r}
# Complex numbers
z <- 3 + 4i
class(z)

# Operations on complex numbers
Mod(z)          # Modulus (5)
Arg(z)          # Argument (angle)
Re(z)           # Real part
Im(z)           # Imaginary part
Conj(z)         # Complex conjugate

# Creating complex numbers
complex(real = 3, imaginary = 4)
```

## Type Coercion

R follows a coercion hierarchy when combining different types:

`logical` → `integer` → `double` → `character`

```{r}
# Implicit coercion
mixed1 <- c(1, 2, TRUE, FALSE)      # All become numeric: 1, 2, 1, 0
mixed1
class(mixed1)

mixed2 <- c(1, 2, "three")          # All become character
mixed2
class(mixed2)

mixed3 <- c(TRUE, 2.5, "text")     # All become character
mixed3
```

### Explicit Coercion

```{r}
# Explicit type conversion
as.logical(c(0, 1, 2, -1))    # FALSE, TRUE, TRUE, TRUE
as.integer(c(1.9, 2.1, 2.9))  # 1, 2, 2 (truncates!)
as.numeric(c(TRUE, FALSE))      # 1, 0

# Coercion may produce NA
as.numeric(c("1", "2", "three"))  # 1, 2, NA
```

::: {.callout-warning}
## Warning
Be careful with `as.integer()` - it truncates (rounds toward zero), not rounds!
Use `round()` first if you need proper rounding.
:::

## Floating Point Precision Issues

### The Problem

Computers use binary floating point representation, which cannot exactly represent many decimal numbers:

```{r}
# Classic example
0.1 + 0.2 == 0.3

# What is the actual value?
print(0.1 + 0.2, digits = 20)
print(0.3, digits = 20)

# Machine epsilon (smallest number that changes 1 when added)
.Machine$double.eps

# Comparing floating point numbers
x <- 0.1 + 0.2
y <- 0.3

# Don't do this
x == y    # FALSE!

# Do this instead
abs(x - y) < 1e-10    # TRUE

# Or use all.equal
all.equal(x, y)        # TRUE
```

### Understanding Machine Precision

```{r}
# Binary representation limitations
for (i in 1:10) {
  print(paste("1/", i, "=", 1/i, "exact:", (i * (1/i)) == 1))
}

# Powers of 2 are exact
1/2 == 0.5      # TRUE
1/4 == 0.25     # TRUE
1/8 == 0.125    # TRUE

# Others are approximate
1/10 == 0.1     # FALSE (in practice, may be TRUE due to rounding)
```

### Best Practices for Floating Point

```{r}
# 1. Use all.equal for comparisons
x <- 1.0000001
y <- 1.0
all.equal(x, y, tolerance = 1e-6)

# 2. Be careful with cumulative operations
sum_rep <- function(n, x) {
  total <- 0
  for (i in 1:n) {
    total <- total + x
  }
  return(total)
}

# Compare
n <- 100000
x <- 0.1
sum_rep(n, x) == n * x    # May be FALSE
all.equal(sum_rep(n, x), n * x)

# 3. Use integer arithmetic when possible
sum_int <- function(n) {
  total <- 0L
  for (i in 1L:n) {
    total <- total + i
  }
  return(total)
}

sum_int(100)
```

## Special Values

### NA (Not Available)

```{r}
# NA represents missing values
x <- c(1, 2, NA, 4, 5)
x

# Propagation of NA
mean(x)              # NA (by default)
mean(x, na.rm = TRUE)  # 3 (ignore NA)

# Checking for NA
is.na(x)             # FALSE FALSE TRUE FALSE FALSE
any(is.na(x))        # TRUE

# Removing NA
x[!is.na(x)]         # 1 2 4 5
na.omit(x)           # removes NA values
```

### NaN (Not a Number)

```{r}
# NaN results from undefined operations
0/0                  # NaN
Inf - Inf            # NaN
sqrt(-1)             # NaN (for real sqrt)

# Checking for NaN
is.nan(NaN)          # TRUE
is.nan(NA)           # FALSE
is.na(NaN)           # TRUE (NaN is also NA)

# NaN propagation
NaN + 5              # NaN
mean(c(1, 2, NaN))   # NaN
```

### Inf and -Inf

```{r}
# Infinity results from overflow
1/0                  # Inf
-1/0                 # -Inf
1e308 * 10           # Inf (overflow)

# Working with infinity
Inf + 1              # Inf
Inf - Inf            # NaN
Inf / Inf            # NaN

# Checking for infinity
is.infinite(Inf)     # TRUE
is.finite(Inf)       # FALSE
is.finite(1)         # TRUE
```

### NULL

```{r}
# NULL represents "nothing" or "undefined"
y <- NULL

# NULL vs NA
length(NULL)         # 0
length(NA)           # 1

# Checking for NULL
is.null(NULL)        # TRUE
is.null(NA)          # FALSE
```

## Type Testing Functions

```{r}
x <- 42L
y <- 42.0
z <- "42"

# Testing types
is.integer(x)        # TRUE
is.numeric(x)        # TRUE (integer is numeric)
is.double(y)         # TRUE
is.character(z)      # TRUE

# Testing special values
is.na(c(1, NA, 3))           # FALSE TRUE FALSE
is.nan(c(1, NaN, 3))         # FALSE TRUE FALSE
is.finite(c(1, Inf, -Inf))   # TRUE FALSE FALSE
is.infinite(c(1, Inf))       # FALSE TRUE
```

## Summary

Key takeaways from this chapter:

1. **Data Types**: R has logical, integer, double, character, complex, and raw atomic types
2. **Coercion**: Types combine following a hierarchy; explicit coercion with `as.*()` functions
3. **Floating Point**: Binary representation cannot exactly store many decimal values
4. **Comparisons**: Use `all.equal()` instead of `==` for floating point comparisons
5. **Special Values**: NA (missing), NaN (undefined), Inf (overflow), NULL (empty)

## Exercises

::: {.callout-important}
## Exercise 1
Explain why `0.1 + 0.2 != 0.3` in R. Write code to properly compare these values.
:::

::: {.callout-important}
## Exercise 2
Create a function `safe_mean(x)` that calculates the mean of a vector, properly handling NA values and returning NA only when all values are NA.
:::

::: {.callout-important}
## Exercise 3
Demonstrate the difference between integer and double arithmetic in R by calculating the sum of 1 to 1 million using both types and comparing results.
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_R_third.pdf`
- R Documentation: `?typeof`, `?is.na`, `?all.equal`
- "The R Inferno" by Patrick Burns
