---
title: "Chapter 2 - Flow Control and Functions in R"
subtitle: "Programming Constructs: Loops, Conditionals, and Custom Functions"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Use if/else statements for conditional execution
- Implement for and while loops
- Write custom functions in R
- Understand variable scope in R
- Apply these concepts to solve practical problems

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_R_second.pdf`
:::

## Conditional Statements

### If Statements

The simplest form of conditional execution:

```{r}
x <- 10

if (x > 5) {
  print("x is greater than 5")
}

# Multiple conditions
y <- 3
if (x > 5 & y < 5) {
  print("Both conditions are true")
}
```

### If-Else Statements

Execute different code blocks based on conditions:

```{r}
score <- 75

if (score >= 90) {
  grade <- "A"
} else if (score >= 80) {
  grade <- "B"
} else if (score >= 70) {
  grade <- "C"
} else if (score >= 60) {
  grade <- "D"
} else {
  grade <- "F"
}

print(paste("Grade:", grade))
```

### The ifelse() Function

Vectorized version for element-wise conditional assignment:

```{r}
ages <- c(15, 22, 17, 30, 16, 45)

# Classify as "adult" or "minor"
status <- ifelse(ages >= 18, "adult", "minor")
status

# Nested ifelse
scores <- c(55, 72, 88, 95, 60, 45)
result <- ifelse(scores >= 90, "Excellent",
            ifelse(scores >= 70, "Good",
            ifelse(scores >= 50, "Pass", "Fail")))
result
```

## Loops in R

### For Loops

Iterate over a sequence:

```{r}
# Basic for loop
for (i in 1:5) {
  print(paste("Iteration:", i))
}

# Loop over vector elements
fruits <- c("apple", "banana", "cherry")
for (fruit in fruits) {
  print(paste("I like", fruit))
}

# Loop with index
values <- c(10, 20, 30, 40, 50)
for (i in seq_along(values)) {
  print(paste("Element", i, "is", values[i]))
}
```

### While Loops

Execute while a condition is true:

```{r}
# Basic while loop
counter <- 1
while (counter <= 5) {
  print(paste("Counter:", counter))
  counter <- counter + 1
}

# Finding when a value exceeds threshold
value <- 1.0
iterations <- 0
while (value < 1000) {
  value <- value * 2
  iterations <- iterations + 1
}
print(paste("Doubled", iterations, "times to reach", value))
```

### Repeat Loops and Break

```{r}
# Repeat with break
x <- 1
repeat {
  print(x)
  x <- x + 1
  if (x > 10) {
    break
  }
}

# Next statement (skip to next iteration)
for (i in 1:10) {
  if (i %% 2 == 0) {
    next  # Skip even numbers
  }
  print(i)  # Only odd numbers printed
}
```

::: {.callout-warning}
## Performance Note
Loops in R are generally slower than vectorized operations. Use vectorization when possible, but loops are fine for complex logic or when readability is more important than speed.
:::

## Functions in R

### Creating Functions

Define reusable code blocks:

```{r}
# Simple function
greet <- function(name) {
  return(paste("Hello,", name, "!"))
}

greet("Alice")
greet("Bob")

# Function with default arguments
power <- function(base, exponent = 2) {
  return(base^exponent)
}

power(3)      # 3^2 = 9 (default exponent)
power(2, 3)   # 2^3 = 8
power(2, exponent = 4)  # Named argument

# Multiple return values (as a list)
stats <- function(x) {
  list(mean = mean(x),
       median = median(x),
       sd = sd(x))
}

data <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
result <- stats(data)
result$mean
result$median
```

### Variable Scope

```{r}
# Global vs local variables
global_var <- 10

my_function <- function() {
  local_var <- 5      # Local variable
  print(global_var)    # Can access global
  return(local_var)
}

my_function()
# local_var is not accessible here
# print(local_var)  # This would cause an error

# Modifying global variable
counter <- 0
increment <- function() {
  counter <<- counter + 1  # Use <<- to modify global
}
increment()
print(counter)  # Now 1
```

## Practical Examples

### Example 1: Calculating π using Buffon's Needle

Buffons needle problem estimates π by dropping needles on a lined surface:

```{r}
buffon_needle <- function(n_trials, needle_length = 1, line_spacing = 1) {
  # Generate random needle positions and angles
  y_center <- runif(n_trials, 0, line_spacing / 2)
  angle <- runif(n_trials, 0, pi / 2)

  # Check if needle crosses a line
  y_tip <- y_center + (needle_length / 2) * sin(angle)
  crosses <- y_tip >= line_spacing / 2

  # Estimate pi
  pi_estimate <- (2 * needle_length * n_trials) / (line_spacing * sum(crosses))
  return(pi_estimate)
}

set.seed(123)
# Run simulation with different numbers of trials
trials <- c(100, 1000, 10000)
for (n in trials) {
  estimate <- buffon_needle(n)
  print(paste("Trials:", n, "- π estimate:", round(estimate, 4)))
}
```

### Example 2: Factorial Function

```{r}
# Recursive factorial
factorial_recursive <- function(n) {
  if (n <= 1) {
    return(1)
  } else {
    return(n * factorial_recursive(n - 1))
  }
}

# Iterative factorial
factorial_iterative <- function(n) {
  result <- 1
  for (i in 2:n) {
    result <- result * i
  }
  return(result)
}

# Compare
factorial_recursive(5)
factorial_iterative(5)
```

### Example 3: Monte Carlo Integration

```{r}
# Estimate integral of f(x) from a to b using Monte Carlo
mc_integrate <- function(f, a, b, n = 10000) {
  # Generate random points
  x <- runif(n, a, b)

  # Evaluate function
  fx <- f(x)

  # Monte Carlo estimate
  integral <- (b - a) * mean(fx)
  return(integral)
}

# Example: integrate x^2 from 0 to 1 (true value = 1/3)
f <- function(x) x^2
estimate <- mc_integrate(f, 0, 1, n = 100000)
print(paste("Monte Carlo estimate:", estimate))
print(paste("True value:", 1/3))
```

### Example 4: Newton-Raphson Method

```{r}
# Find root of f(x) = 0 using Newton-Raphson
newton_raphson <- function(f, df, x0, tolerance = 1e-6, max_iter = 100) {
  x <- x0
  for (i in 1:max_iter) {
    fx <- f(x)
    if (abs(fx) < tolerance) {
      return(list(root = x, iterations = i))
    }
    dfx <- df(x)
    if (abs(dfx) < 1e-10) {
      stop("Derivative too small")
    }
    x <- x - fx / dfx
  }
  stop("Maximum iterations reached")
}

# Example: Find sqrt(2) by solving x^2 - 2 = 0
f <- function(x) x^2 - 2
df <- function(x) 2 * x

result <- newton_raphson(f, df, x0 = 1)
print(paste("Square root of 2:", result$root))
print(paste("Iterations:", result$iterations))
```

## Summary

Key takeaways from this chapter:

1. **Conditionals**: Use `if`, `else if`, `else` for branching; `ifelse()` for vectorized conditionals
2. **Loops**: `for` for fixed iterations, `while` for condition-based, `repeat` with `break` for flexibility
3. **Functions**: Define with `function()`, use named arguments and defaults for clarity
4. **Scope**: Variables are local by default; use `<<-` to modify global variables (use sparingly)
5. **Performance**: Prefer vectorization over loops when possible

## Exercises

::: {.callout-important}
## Exercise 1
Write a function `is_prime(n)` that returns `TRUE` if n is prime and `FALSE` otherwise. Test it on numbers from 1 to 20.
:::

::: {.callout-important}
## Exercise 2
Implement the Fibonacci sequence using both a loop and recursion. Compare the execution time for calculating the 30th Fibonacci number.
:::

::: {.callout-important}
## Exercise 3
Create a function that approximates the integral of sin(x) from 0 to π using the trapezoid rule with n subdivisions. Test with n = 10, 100, 1000.
:::

::: {.callout-important}
## Exercise 4
Write a simulation to estimate the probability of getting at least one 6 when rolling a die 6 times.
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_R_second.pdf`
- R Documentation: `?Control`, `?function`
- "Advanced R" by Hadley Wickham
