---
title: "Chapter 9 - Bayesian Parameter Estimation"
subtitle: "Conjugate Priors, Posterior Analysis, and Point Estimation"
number-sections: true
---

## Learning Objectives

After completing this chapter, you should be able to:

- Understand the concept of conjugate priors
- Compute posterior distributions for common models
- Calculate posterior summary statistics
- Apply MAP and LMS estimators
- Understand the reliability of Bayesian estimates

## Overview

::: {.callout-note}
This chapter covers material from the lecture slides: `Advanced_Statistics_ParameterEstimation.pdf`
:::

## Bayes' Theorem and Bayesian Inference

### The Framework

Bayesian inference updates our beliefs about parameters $\theta$ after observing data $D$:

$$P(\theta | D, H) = \frac{P(D | \theta, H) P(\theta | H)}{P(D | H)}$$

where:
- $P(\theta | H)$ = **Prior** (initial belief)
- $P(D | \theta, H)$ = **Likelihood** (probability of data given parameters)
- $P(\theta | D, H)$ = **Posterior** (updated belief)
- $P(D | H)$ = **Evidence** (marginal likelihood, normalizing constant)

## Conjugate Priors

A prior is **conjugate** to a likelihood if the posterior has the same form as the prior.

### Beta-Bernoulli / Beta-Binomial

For estimating a probability $p$ from $k$ successes in $n$ trials:

**Prior**: $p \sim \text{Beta}(\alpha, \beta)$

**Likelihood**: $k | p \sim \text{Binomial}(n, p)$

**Posterior**: $p | k \sim \text{Beta}(\alpha + k, \beta + n - k)$

### Posterior Derivation

```{r}
# Example: Coin flipping
# Prior: Beta(2, 2) - slightly biased toward fairness
# Data: 7 heads out of 10 flips

alpha_prior <- 2
beta_prior <- 2
k <- 7  # heads
n <- 10  # total flips

# Posterior
alpha_post <- alpha_prior + k
beta_post <- beta_prior + n - k

cat("Posterior: Beta(", alpha_post, ",", beta_post, ")\n")

# Plot prior and posterior
p <- seq(0, 1, length.out = 200)
prior <- dbeta(p, alpha_prior, beta_prior)
posterior <- dbeta(p, alpha_post, beta_post)

plot(p, prior, type = "l", col = "blue", lty = 2,
     ylab = "Density", xlab = "p",
     main = "Beta-Binomial: Prior and Posterior")
lines(p, posterior, col = "red", lwd = 2)
abline(v = k/n, col = "green", lty = 3)  # MLE (sample proportion)
legend("topright", legend = c("Prior", "Posterior", "MLE"),
       col = c("blue", "red", "green"), lty = c(2, 1, 3), lwd = c(1, 2, 1))
```

### Effect of Sample Size

```{r}
# Show how posterior concentrates as n increases
p <- seq(0, 1, length.out = 1000)
true_p <- 0.7
n_values <- c(10, 50, 200)

# Uniform prior
alpha <- 1
beta <- 1

colors <- c("red", "blue", "darkgreen")
plot(0, 0, xlim = c(0, 1), ylim = c(0, 15), type = "n",
     xlab = "p", ylab = "Density",
     main = "Posterior Concentration with Increasing Data")

for (i in seq_along(n_values)) {
  n <- n_values[i]
  k <- round(true_p * n)
  posterior <- dbeta(p, alpha + k, beta + n - k)
  lines(p, posterior, col = colors[i], lwd = 2)
}
abline(v = true_p, col = "black", lty = 2)
legend("topright", legend = c(paste0("n=", n_values), "True p"),
       col = c(colors, "black"), lty = c(1, 1, 1, 2), lwd = c(2, 2, 2, 1))
```

## Posterior Summary Statistics

### Posterior Mean (LMS - Least Mean Square)

For Beta$(\alpha, \beta)$:

$$\mathbb{E}[p | k, n] = \frac{\alpha + k}{\alpha + \beta + n}$$

As $n \to \infty$: $\mathbb{E}[p | \text{data}] \to \frac{k}{n}$ (empirical frequency)

### Posterior Mode (MAP - Maximum A Posteriori)

$$p_{\text{MAP}} = \frac{\alpha + k - 1}{\alpha + \beta + n - 2}$$

(Valid when $\alpha + k > 1$ and $\beta + n - k > 1$)

### Posterior Variance

$$\text{Var}(p | k, n) = \frac{(\alpha + k)(\beta + n - k)}{(\alpha + \beta + n)^2(\alpha + \beta + n + 1)}$$

As $n \to \infty$: variance $\to 0$ (posterior concentrates)

### In R

```{r}
# Calculate posterior statistics
alpha <- 7   # from Beta(7, 5) posterior
beta <- 5

# Posterior mean
post_mean <- alpha / (alpha + beta)
cat("Posterior mean:", post_mean, "\n")

# Posterior mode
post_mode <- (alpha - 1) / (alpha + beta - 2)
cat("Posterior mode (MAP):", post_mode, "\n")

# Posterior variance
post_var <- alpha * beta / ((alpha + beta)^2 * (alpha + beta + 1))
cat("Posterior variance:", post_var, "\n")
cat("Posterior std:", sqrt(post_var), "\n")
```

## Point Estimation

### General Framework

We often want a single estimate $\hat{\theta}$ to quote from the posterior.

The optimal estimate depends on the **loss function** $L(a, \theta)$:

$$\hat{\theta}(X) = \arg\min_a \mathbb{E}[L(a, \theta) | X]$$

### MAP Estimator (0-1 Loss)

For 0-1 loss: $L(a, \theta) = \begin{cases} 0 & a = \theta \\ 1 & a \neq \theta \end{cases}$

$$\hat{\theta}_{\text{MAP}} = \arg\max_\theta p(\theta | X)$$

**Pros**:
- Simple, computationally convenient
- Natural interpretation (most probable value)

**Cons**:
- Uses only local information (mode may not be representative)
- Not invariant under reparameterization

### LMS Estimator (Squared Loss)

For squared loss: $L(a, \theta) = (a - \theta)^2$

$$\hat{\theta}_{\text{LMS}} = \mathbb{E}[\theta | X]$$

**Pros**:
- Uses full posterior distribution
- Invariant under reparameterization

**Cons**:
- Sensitive to tails
- May not be the "most likely" value

### Comparison

```{r}
# Example where MAP and mean differ
# Bimodal posterior
p <- seq(0, 1, length.out = 500)
posterior <- 0.6 * dbeta(p, 2, 8) + 0.4 * dbeta(p, 8, 2)

# Normalize
posterior <- posterior / sum(posterior) * length(p)

# Find MAP
map_idx <- which.max(posterior)
map_estimate <- p[map_idx]

# Approximate mean
mean_estimate <- sum(p * posterior) / sum(posterior)

plot(p, posterior, type = "l", ylab = "Density",
     main = "Bimodal Posterior: MAP vs Mean")
abline(v = map_estimate, col = "red", lty = 2, lwd = 2)
abline(v = mean_estimate, col = "blue", lty = 3, lwd = 2)
legend("topright", legend = c("Posterior", "MAP", "Mean"),
       col = c("black", "red", "blue"), lty = c(1, 2, 3), lwd = c(1, 2, 2))
```

## Gaussian Approximation

For large samples, the posterior can be approximated by a Gaussian centered at the MAP.

### Laplace Approximation

Taylor expansion of log-posterior around MAP:

$$\mathcal{L}(\theta) \approx \mathcal{L}(\theta_{\text{MAP}}) - \frac{1}{2}(\theta - \theta_{\text{MAP}})^T H (\theta - \theta_{\text{MAP}})$$

where $H$ is the negative Hessian matrix of log-posterior.

Posterior $\approx \mathcal{N}(\theta_{\text{MAP}}, C)$ where $C = H^{-1}$

### Final Estimate

$$\hat{\theta} = \theta_{\text{MAP}} \pm \sqrt{\text{diag}(C)}$$

### Bernoulli Example

```{r}
# Gaussian approximation for Beta posterior
alpha <- 12
beta <- 8
n <- alpha + beta - 2  # effective sample size
k <- alpha - 1

# MAP
map <- (alpha - 1) / (alpha + beta - 2)

# Approximate variance (inverse Fisher information)
# For Bernoulli: var ≈ p(1-p)/n
approx_var <- map * (1 - map) / n

p <- seq(0, 1, length.out = 200)
exact <- dbeta(p, alpha, beta)
gaussian_approx <- dnorm(p, mean = map, sd = sqrt(approx_var))

plot(p, exact, type = "l", ylab = "Density",
     main = "Beta Posterior vs Gaussian Approximation")
lines(p, gaussian_approx, col = "red", lty = 2)
legend("topright", legend = c("Exact (Beta)", "Gaussian Approx"),
       col = c("black", "red"), lty = c(1, 2))
```

## Loss and Risk

### Posterior Risk

The expected loss given the data:

$$\rho(a | x) = \int L(a, \theta) p(\theta | x) d\theta$$

### Bayes Risk

The overall quality of an estimator:

$$\mathcal{R} = \int \rho(\hat{\theta}(x) | x) p(x) dx$$

### Key Result

Minimizing the posterior risk for each $x$ automatically minimizes the global Bayes risk.

**Local optimum = Global optimum**

## Take-Home Messages

1. **Bayesian estimation minimizes expected loss given the data**
2. **The local decision rule is already globally optimal**
3. **Conjugate priors make computation tractable**
4. **Posterior concentrates as data grows (consistency)**
5. **MAP is simple but not always representative; Mean uses full posterior**

## Summary

| Estimator | Loss Function | Formula | Properties |
|-----------|--------------|---------|------------|
| MAP | 0-1 | $\arg\max p(\theta | x)$ | Simple, not invariant |
| LMS | Squared | $\mathbb{E}[\theta | x]$ | Uses full posterior, invariant |

Key takeaways from this chapter:

1. **Conjugate priors**: Prior and posterior have same form (Beta-Binomial, Gamma-Poisson, etc.)
2. **MAP**: Maximum a posteriori - simplest point estimate
3. **LMS**: Posterior mean - uses full distribution
4. **Gaussian approximation**: For large samples, posterior ≈ Gaussian centered at MAP
5. **Loss functions**: Define what "good" estimate means; squared loss gives mean, 0-1 gives mode

## Exercises

::: {.callout-important}
## Exercise 1
You have a Beta(3, 3) prior for a coin probability and observe 15 heads in 20 flips. Calculate the posterior distribution, posterior mean, and posterior variance.
:::

::: {.callout-important}
## Exercise 2
Explain why MAP is not invariant under reparameterization. Give a concrete example using the logit transformation $\varphi = \log\frac{p}{1-p}$.
:::

::: {.callout-important}
## Exercise 3
Show that for a Bernoulli likelihood with Beta prior, the posterior variance decreases as $O(1/n)$ where $n$ is the sample size.
:::

## Further Reading

- **Original Slides**: `../Advanced_Statistics_ParameterEstimation.pdf`
- "Bayesian Data Analysis" by Gelman et al.
- "Pattern Recognition and Machine Learning" by Bishop
