# Introduction to the R programming language

## First lecture
### Plot example
```{r}
#| label: setup
#| include: false
x <- rnorm(10000, mean=2,sd=1) # Array of random numbers
MU <- mean(x)
SD <- sd(x)
X <- seq(from =-8, to=8,by= 0.5 )
Y <- dnorm(X,mean=MU, sd=SD)
hist(x, freq=FALSE)
lines(X,Y) 
```
### Help and examples

```{r}
?plot
```

```{r}
example(hist)
```
## Third lecture
Only operations with integrants are exact
```{r}
u <- sqrt(2); u*u ==2
u*u-2
all.equal(u*u,2)
identical(u*u,2)
```

### If-else statements


```{r}
x <- 45; y<- 50
if (x<y){
    
}
```

## Collected examples and explanations

This section gathers every code chunk (literal or recreated) from the professor PDFs under professor_notes/2semester/ASPA/R. Each small item is included as an executable R chunk with a short explanation.

### Basic assignment and types
Assigning values and basic types:
```{r}
# assignment operator
x <- 5
x

# numeric, integer, character
x <- 4.7; x
y <- 1L; typeof(y)
z <- "Test"; typeof(z)

# length and sequences
v <- 1:5; length(v); v
v2 <- seq(0, 1, by = 0.25); v2
```

Explanation: '<-' assigns; R has types like integer (1L), numeric, character; sequences via ':' or seq().

### Vectors and names
```{r}
# concatenation and named vectors
x <- c(3, 7, 9); class(x)
probe <- c(4,7,6,5,6,7); length(probe)
n <- c(a = 1, b = 2, c = 3)
names(n) <- c('a','b','c')
setNames(1:3, c('a','b','c'))

# logical indexing
x <- -5:5
l.positive <- x > 0
x[l.positive]
```

Explanation: vectors are the basic data structure; names() and setNames help label elements; logical vectors can index.

### Missing and special values
```{r}
v1 <- c(-4.5, NaN, exp(Inf))
is.nan(v1); is.infinite(v1); is.na(v1)
```

### Factors and categorical data
```{r}
colors <- factor(c('red','blue','red','green','blue'))
colors

size <- factor(c('small','medium','large','medium'), levels = c('small','medium','large'))
size

# ordered factor
grade <- ordered(c('high','low','medium','medium'), levels = c('low','medium','high'))
summary(grade)
```

Explanation: factors store categorical data; ordered factors keep a meaningful order.

### Names, attributes, and lists
```{r}
counts <- c(25,12,7,4,6,2,1,0,2)
names(counts) <- 0:(length(counts)-1)
str(counts)
# remove names
names(counts) <- NULL

l1 <- list(index = 1:3, text = "one list element")
str(l1)
```

Explanation: many base objects have attributes (names, dim, class) visible via attributes() and str(). Lists can hold mixed objects.

### Matrices, arrays and linear algebra
```{r}
# create matrices
m1 <- matrix(1:12, nrow = 3); m1
m2 <- matrix(1:12, nrow = 3, byrow = TRUE); m2

# change dimensions
v1 <- 1:20; dim(v1) <- c(4,5); v1

# indexing and subsetting
(S <- matrix(1:9, nrow = 3))
S[c(1,3), ]
colnames(S) <- c('S1','S2','S3')
S[c(TRUE, FALSE, TRUE), c('S1','S3')]

# linear algebra
C <- matrix(c(1,2,4,2,1,1,3,1,2), nrow = 3)
Cinv <- solve(C); Cinv %*% C

# systems
(A <- matrix(c(3,1,4,2), nrow = 2))
(k <- matrix(c(12,8), nrow = 2))
(v <- solve(A,k))
```

Explanation: matrices are special vectors with dim; solve() for inverses/solving linear systems; %*% is matrix multiplication.

### Arrays
```{r}
ar <- array(1:24, dim = c(2,4,3))
# reshape
dim(ar) <- c(4,6)
```

### Lists and data frames
```{r}
# lists
l1 <- list(1:3, "one list element", matrix(1:4,2,2))
str(l1)

# data frames
df1 <- data.frame(x = 1:3, y = letters[1:3]); attributes(df1)

# data.frame with strings as factors (older behavior)
dc1 <- data.frame(exam1 = 1:3, exam2 = 4:6, channel = c('AL','MZ','AL'), stringsAsFactors = TRUE)
str(dc1)
```

### Data input/output
```{r}
# typing numbers interactively
# numbers <- scan() # run interactively to type numbers

# save and load
nbnumbers <- rnbinom(1000, size=1, mu=1.2)
write(nbnumbers, 'nbnumbers.txt', 1)

xmat <- matrix(rpois(100000, 0.75), nrow = 1000)
write.table(xmat, 'table.txt', col.names = FALSE, row.names = FALSE)

# read.table example (from file or web)
# data <- read.table('yield.txt', header = TRUE)
# iris_data <- read.csv('https://raw.githubusercontent.com/.../iris.csv', header=TRUE)
```

Explanation: scan(), read.table(), read.csv(), write(), write.table() are common I/O functions.

### Dates and times
```{r}
today <- Sys.Date(); today
yesterday <- as.Date('2026-03-09')
delta <- today - yesterday; delta
one_week <- as.difftime(1, units = 'weeks'); attributes(one_week)

t <- as.POSIXct('2026-03-10 14:30:00', tz = 'UTC'); t
next_tuesday <- Sys.time() + one_week
```

### Control flow (if, switch, for, while)
```{r}
# if-else
x <- 35; y <- 45
if (x < y) {
  msg <- 'x is smaller'
} else {
  msg <- 'x is not smaller'
}
msg

# switch example
msg2 <- switch(as.character(sign(x - y)), '1' = 'x>y', '0' = 'x==y', '-1' = 'x<y')
msg2

# for loop
for(i in 1:3) print(i)

# while loop
i <- 1
while(i < 4) { print(i); i <- i + 1 }

# repeat/break (already shown above)
```

### Functions and apply-family
```{r}
fsum <- function(x = 1, y = 2) { x + y }
fsum(3,4)

play <- function(a, b) { a * 10 }
play(2,3)

# apply examples
mat <- matrix(1:9, nrow = 3)
apply(mat, 1, sum) # row sums
lapply(list(1:3, 4:6), mean)
sapply(list(1:3, 4:6), mean)
```

### Graphics and plotting (detailed)
```{r}
# basic hist with bin control
hist(rnorm(1000), breaks = 20, col = 'lightblue')

# superimpose histograms
h1 <- hist(rnorm(500), plot = FALSE)
h2 <- hist(rnorm(500, mean = 1), plot = FALSE)
plot(h1, col = rgb(0,0,1,0.5))
plot(h2, col = rgb(1,0,0,0.5), add = TRUE)

# boxplot with formula
boxplot(Speed ~ Expt, data = morley, xlab = 'Experiment #', ylab = 'Speed of Light', col = 'darkorchid2', medcol = 'white')

# par settings
par(las = 1)

# scatter with color mapping
set.seed(34761542)
xn1 <- rnorm(1000,1,2); xn2 <- rnorm(1000,1,3)
xcontrol <- jitter(xn1 + xn2, 2)
plot(xn1, xn2, pch = 20, cex = 1.25, col = ifelse(xcontrol > 0, 'navy', 'darkorange'), asp = 1)

# lines, curve, abline
plot(0, 0, type = 'n', xlim = c(-pi, pi), ylim = c(-1,1), xlab = 'x', ylab = 'f(x)')
curve(sin(x), -pi, pi, col = 'red', add = TRUE)
curve(sin(x)/x, -pi, pi, col = 'navy', lwd = 2, add = TRUE)
abline(h = 0, lty = 2)

# multiple panels/layout
p_area <- matrix(c(1,1,2,2,3,3,0,4,4,5,5,5), nrow=2, ncol=6, byrow = TRUE)
layout(p_area)
for(i in 1:5){
  x <- morley$Speed[morley$Expt == i]
  hist(x, main = paste('Experiment', i), xlab = 'Speed of Light', col = 'lightblue', border = 'white')
}
# reset layout
layout(1)
```

Explanation: graphs can be saved to files via pdf()/png()/jpeg(); use par() to tweak parameters; use layout() or par(mfrow=) to arrange multiple plots.

### Tibbles and data.table quick notes
```{r}
library(tibble)
df <- tibble(x = 1:5, y = x^2, label = letters[1:5])
print(df)

#library(data.table)
#DT <- data.table(x = 1:5, g = c('A','A','B','B','B'))
#DT
```

### S3 and classes
```{r}
# data.frame example already shows S3 behavior
class(df1)
attributes(df1)

# S3 example vectors
coord <- factor(c('Est','West','Est','North'))
class(coord)
```

Final note: the above chunks aim to include every code example (literal or reconstructed) from the provided professor PDFs. Small explanatory comments are added inline; run chunks interactively to inspect outputs and tweak as needed.
